---
title: "Some Mapping Options in R (More Ways to Skin the Cat)"
author: "Daniel Palacios, OSU Fisheries & Wildlife and Marine Mammal Institute"
date: "HMSC UseR Meeting, February 10, 2015"
output: ioslides_presentation
---

## Intro

The presentation at the Jan. 2015 meeting by Tom Wainwright:    

"Skinning Cats with R: Bathymetric & Coastline Maps"    
http://rpubs.com/hmsc_useR/BathyMapDemo     

covered aspects of mapping in R. Today's presentation covers additional 
approaches.


## Scientific Mapping (+ Geospatial Analysis)

* GIS
    + ESRI (lots of $)
    + QGIS (free)
* GMT: Generic Mapping Tools (free)
* Matlab
    + Mapping Toolbox (lots of $)
    + M_Map (free)
* R (free)   
    <div class="red"> ... welcome to the jungle? <div/>


## R Packages (not exhaustive)

* Mapping:
    + *rworldmap*, *maps*, *maptools*, *PBSmapping*, *marmap*,    
    *RgoogleMaps*, *OpenStreetMap*, *ggplot2*, *ggmap* ...
* Spatial analysis:
    + *raster*, *sp (spplot)*, *rgdal*, *rgeos*, *spatstat*, ...
* Related data:
    + *rworldmap (rworldxtra)*, *marmap*, ...
    + *ggmap (Google Maps, OpenStreetMaps, Stamen Maps, CloudMade)*


## External Data

> * Bathymetry & shorelines:
    + NOAA National Geographic Data Center  
    http://www.ngdc.noaa.gov/mgg/bathymetry/relief.html
    http://www.ngdc.noaa.gov/mgg/shorelines/gshhs.html
    + SRTM30_Plus ("Smith & Sandwell")
    http://topex.ucsd.edu/WWW_html/srtm30_plus.html
    + GSHHG - A Global Self-consistent, Hierarchical,   
    High-resolution Geography Database
    http://www.soest.hawaii.edu/pwessel/gshhg/

> * Natural Earth (beautiful background tiles & more data!)   
    http://www.naturalearthdata.com


## External Data

> * Environmental data:
    + NOAA's ERD Data Access Protocols (ERDDAPs)  
    http://coastwatch.pfel.noaa.gov/erddap/index.html
    http://upwell.pfeg.noaa.gov/erddap/index.html
    + NASA (JPL, GSFC)    
    http://podaac.jpl.nasa.gov
    http://oceancolor.gsfc.nasa.gov/cms/
    http://giovanni.gsfc.nasa.gov/giovanni/
    + AVISO http://www.aviso.altimetry.fr/en/home.html

> * ERRDAP's xtract-o-matic routines for single points, animal tracks, or 
    irregularly shaped polygons:   
    http://coastwatch.pfel.noaa.gov/xtracto/

> * Your favorite shapefile repositories


## Mapping
### Two approaches:
1. Draw all your data (including geographic data)
    * Standard format for scientific publications

2. Overlay your data on a georeferenced tile (pretty!)
    * Appropriate for slide and poster presentations, magazines, etc.

## Mapping
### Basic example:
Plot the location of 5 sampling stations along the central California coast, 
USA, using some of the mapping options available in R.

```{r}
# Specify station locations:
coords.locs <- data.frame(Locs = c("Pt. Piedras Blancas",
                                   "Pt. Estero","Pt. Buchon",
                                   "Pt. Sal", "Pt. Arguello"),
                          Long = c(-121.3500, -120.8000, -120.7000,
                                   -120.6000, -120.5500),
                          Lat = c(35.6500, 35.4000, 35.2000,
                                  34.8000, 34.5500))

# Specify desired map boundaries:
xlim <- c(-122.3333, -120.5)
ylim <- c(34.3333, 36.3333)
```


## Mapping
### Load libraries required for the example:

```{r, eval = FALSE}
library(maps)
library(mapdata) # loads additional map databases (high-res)
library(maptools) # (loads foreign, sp, grid, and lattice)
library(rgeos) # required by maptools to work with GHHSG
#gpclibPermit() # older alternative to work with GHHSG
library(PBSmapping)
library(ggplot2)
library(ggmap)
```

```{r, echo = FALSE}
library(maps)
library(mapdata) # loads additional map databases (high-res)
library(maptools) # (loads foreign, sp, grid, and lattice)
library(rgeos) # required by maptools to work with GHHSG
#gpclibPermit() # older alternative to work with GHHSG
library(PBSmapping)
library(ggplot2)
library(ggmap)
```


## Mapping

- Some packages can use high-resolution coastlines
- GSHHG (formerly GSHHS): http://www.soest.hawaii.edu/pwessel/gshhg/    
- Download the BINARY file
- The ZIP file (~120 MB) HAS to be extracted into one of R's    
packages (e.g., maptools)... on my Mac machine:    

"/Library/Frameworks/R.framework/Versions/3.3/Resources/library/
maptools/share/gshhg-bin-2.3.5"

```{r}
# Call the GSHHG file with the highest-resolution coastline:
callGSHHG <- system.file("share/gshhg-bin-2.3.5/gshhs_h.b",
                         package = "maptools", mustWork = TRUE)
```


## 1a. Drawing: *maps* package

```{r}
# Generate map object (uses maps coarse coastline):
coastCAc <- map(database = "state", regions = "california",
                plot = FALSE)
```

- *coastCAc* is a map class object with four variables:     
        *x*, *y*, *range*, and *names*   
- If projection is not specified a "rectangular" projection     
  is used
- [Note: package *mapdata* offers higher resolution coastlines 
that may work with examples 1a and 1d, but I haven't checked]


## 1a. Drawing: *maps* package

```{r, eval = FALSE}
# Plot the base map:
map(database = "state", regions = "california", fill = TRUE,
    col = "grey85", xlim = xlim, ylim = ylim)
# map(coastCAc) and plot(coastCAc) are also options
map.axes() # adds the axes, as these are not drawn by default

# Add the locations along with the labels:
points(coords.locs[,2:3], pch = 20, col = "red")
text(coords.locs[,2]-0.1,coords.locs[,3], coords.locs[,1],
     adj = 1, cex = 0.6)
grid(lty = 1)
box()
```


## 1a. Drawing: *maps* package

Plain and simple. Coastline is a bit coarse. Use GSHHG!

```{r, echo = FALSE}
# Plot the base map:
map(database = "state", regions = "california", fill = TRUE,
    col = "grey85", xlim = xlim, ylim = ylim)
# map(coastCAc) and plot(coastCAc) are also options
map.axes() # adds the axes, as these are not drawn by default

# Add the locations along with the labels:
points(coords.locs[,2:3], pch = 20, col = "red")
text(coords.locs[,2]-0.1,coords.locs[,3], coords.locs[,1],
     adj = 1, cex = 0.6)
grid(lty = 1)
box()
```


## 1b. Drawing: *maptools* package (uses GSHHG)

```{r}
# Extract high-resolution coastline with long in 180-deg system:
coastCAh180 <- Rgshhs(callGSHHG, xlim = xlim+360, ylim = ylim,
                      level = 1, shift = TRUE)
```


## 1b. Drawing: *maptools* package (uses GSHHG)

```{r, eval = FALSE}
# Plot the base map using GSHHG high-resolution coastline:
plot(coastCAh180$SP, col = "grey85", pbg = "azure",
     xlim = xlim, ylim = ylim, xaxs = "i", yaxs = "i",
     axes = TRUE)

# Add the locations along with the labels:
points(coords.locs[,2:3], pch = 20, col = "red")
text(coords.locs[,2]-0.1,coords.locs[,3], coords.locs[,1],
     adj = 1, cex = 0.6)
grid(lty = 1)
box()
```


## 1b. Drawing: *maptools* package (uses GSHHG)

Locations on land (inaccurate coords.locs?). x-axis too wide. Blue background 
argument (*pbg*) did not work.

```{r, echo = FALSE}
# Plot the base map using GSHHG high-resolution coastline:
plot(coastCAh180$SP, col = "grey85", pbg = "azure",
     xlim = xlim, ylim = ylim, xaxs = "i", yaxs = "i",
     axes = TRUE)

# Add the locations along with the labels:
points(coords.locs[,2:3], pch = 20, col = "red")
text(coords.locs[,2]-0.1,coords.locs[,3], coords.locs[,1],
     adj = 1, cex = 0.6)
grid(lty = 1)
box()
```


## 1c. Drawing: *PBSmapping* package (uses GSHHG)

```{r}
# Extract high-resolution coastline with long in 360-deg system:
coastCAh360 <- importGSHHS(callGSHHG, xlim = xlim+360,
                           ylim = ylim, maxLevel = 1)
```


## 1c. Drawing: *PBSmapping* package (uses GSHHG)

```{r}
# Specify the location coordinates as PBSmapping Event data type:
coords.locs2 <- cbind(Nbr = 1:5, coords.locs[, 2:3])
names(coords.locs2) <- c("EID", "X", "Y")
coords.locs2 <- as.EventData(coords.locs2)
```


## 1c. Drawing: *PBSmapping* package (uses GSHHG)

```{r, eval = FALSE}
# Plot the base map:
par(mgp = c(2, 0, 0)) # shift up the line where the x-axis legend goes
                      # Defaults are mgp = c(3,1,0)
plotMap(coastCAh360, col = "grey85", bg = "lightblue")

# Add the locations along with the labels:
addPoints(coords.locs2, pch = 20, col = "red")
text(coords.locs[, 2]-0.05, coords.locs[, 3], coords.locs[, 1],
     adj = 1, cex = 0.6)
```


## 1c. Drawing: *PBSmapping* package (uses GSHHG)

Looks good. Nice fill options for land and ocean. Ticks are inside plot area.
```{r, echo = FALSE}
# Plot the base map:
par(mgp = c(2, 0, 0)) # shift up the line where the x-axis legend goes
# Defaults are mgp = c(3, 1, 0)
plotMap(coastCAh360, col = "grey85", bg = "lightblue")

# Add the locations along with the labels:
addPoints(coords.locs2, pch = 20, col = "red")
text(coords.locs[, 2]-0.05, coords.locs[, 3], coords.locs[, 1],
     adj = 1, cex = 0.6)
```


## 1d. Drawing: *ggplot2* package (uses *maps* package)

```{r}
# Use state coastline from maps package. Notice call differs
# slightly from call in slide 9)
coastCA3 <- data.frame(map(database = "state",
                           regions = "california",
                           plot = FALSE)[c("x", "y")])
# has to be "x / y"!
```


[Note: It may be that package *mapdata* offers higher resolution coastlines 
for use with *maps* (ex. 1a and 1d), but I haven't checked]


## 1d. Drawing: *ggplot2* package (uses *maps* package)

```{r, eval = FALSE}
# Generate full map by layering:
ggplot(coastCA3, aes(x = x, y = y)) + geom_path() +
  coord_map(project = "mercator", xlim = xlim, ylim = ylim) +
  geom_point(data = coords.locs, aes(Long, Lat),
             size = 4, color = "red") +
  geom_point(data = coords.locs[1:3,], aes(Long, Lat),
             size = 2, color = "black") +
  geom_text(data = coords.locs, aes(x = Long-0.05, y = Lat,
                                    label = Locs),
            hjust = 1, size = 3, color = "black") +
  labs(x = "Longitude", y = "Latitude", size = 20) +
  theme(axis.text = element_text(size = rel(1.25)),
        axis.title = element_text(size = rel(1.25))) +
  theme_bw()
```


## 1d. Drawing: *ggplot2* package (uses *maps* package)

Coarse coastline. No land fill options with *geom_path()*. Crisp plot area.
```{r, echo = FALSE}
# Generate full map by layering:
ggplot(coastCA3, aes(x = x, y = y)) + geom_path() +
  coord_map(project = "mercator", xlim = xlim, ylim = ylim) +
  geom_point(data = coords.locs, aes(Long, Lat),
             size = 4, color = "red") +
  geom_point(data = coords.locs[1:3, ], aes(Long, Lat),
             size = 2, color = "black") +
  geom_text(data = coords.locs, aes(x = Long-0.05, y = Lat,
                                    label = Locs),
            hjust = 1, size = 3, color = "black") +
  labs(x = "Longitude", y = "Latitude", size = 20) +
  theme(axis.text = element_text(size = rel(1.25)),
        axis.title = element_text(size = rel(1.25))) +
  theme_bw()
```


## 2a. Overlay on tile: *ggplot2* package (uses *OpenStreetMap* package)

```{r, eval = FALSE}
# Load additional libraries:
library(RColorBrewer) # nice color schemes
Sys.setenv(NOAWT = 1) # call this before loading OpenStreetMap!
library(OpenStreetMap) # base-layer maps
```

```{r, echo = FALSE}
# Load additional libraries:
Sys.setenv(NOAWT = 1) # call this before loading OpenStreetMap!
library(OpenStreetMap) # base-layer maps
library(RColorBrewer) # nice color schemes
```


## 2a. Overlay on tile: *ggplot2* package (uses *OpenStreetMap* package)
```{r}
# For plotting with OpenStreetMap, coordinates need to be converted
# from long/lat to Mercator using projectMercator().
coords.locs2 <- cbind(coords.locs, projectMercator(coords.locs$Lat,
                                            coords.locs$Long))
# This adds new columns 'x' and 'y' to coords.locs2

# Download base-layer map from the web:
base.lyr <- openmap(c(ylim[2], xlim[1]), c(ylim[1], xlim[2]),
                    type = "bing")

# Create base map layer:
base.map = autoplot(base.lyr)
```


## 2a. Overlay on tile: *ggplot2* package (uses *OpenStreetMap* package)

```{r, eval = FALSE}
# Build up map by adding elements as layers:
prettyMap <- base.map +
  geom_point(aes(x = x, y = y), size = 4.5, color = "white",
             data = coords.locs2) +
  geom_point(aes(x = x, y = y, size = 4, color = LETTERS[1:5]),
             data = coords.locs2) +
  scale_colour_brewer(palette = "Set2") +
  geom_text(aes(label = Locs, x = x-8000, y = y), color = "white",
            hjust = 1, data = coords.locs2) +
  guides(color = guide_legend("Station Code"), size = FALSE) +
  labs(title = "Station Locations \nOct. 2014 - Feb. 2015",
       x = "Longitude", y = "Latitude") +
  theme(axis.text = element_text(size = rel(1)))

# Print to graphics device:
print(prettyMap)
```


## 2a. Overlay on tile: *ggplot2* package (uses *OpenStreetMap* package)

[Note: package *ggmap* achieves similar results]

```{r, echo = FALSE}
# Build up map by adding elements as layers:
prettyMap <- base.map +
  geom_point(aes(x = x, y = y), size = 4.5, color = "white",
             data = coords.locs2) +
  geom_point(aes(x = x, y = y, size = 4, color = LETTERS[1:5]),
             data = coords.locs2) +
  scale_colour_brewer(palette = "Set2") +
  geom_text(aes(label = Locs, x = x-8000, y = y), color = "white",
            size = 3, hjust = 1, data = coords.locs2) +
  guides(color = guide_legend("Station Code"), size = FALSE) +
  labs(title = "Station Locations \nOct. 2014 - Feb. 2015",
       x = "Longitude", y = "Latitude") +
  theme(axis.text = element_text(size = rel(1)))

# Print to graphics device:
print(prettyMap)
```
